---
title: "Lab 3"
author: "Gustav Sternelöv"
date: "3 maj 2016"
output: pdf_document
---

# Assignment 1 - Normal model, mixture of normal model with semi-conjugate prior.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#### Assignment 1 ####
rainfall <- read.delim("C:/Users/Gustav/Documents/Bayesian-Learning/Lab 3/rainfall.dat", sep="", header = TRUE)
library(ggplot2)
library(gridExtra)
```
## a) Normal model.

### i)
```{r,echo=FALSE}
## a)
# priors and others
mu0 <- 1
kappa0 <- 1
v0 <- 1
sigma0 <- 1
n <- nrow(rainfall)
ybar <- colMeans(rainfall)
s2 <- var(rainfall[,1])

# Posteriors
muN <- (kappa0 / (kappa0 + n)) * mu0 + (n / (kappa0 + n)) * ybar 
kappaN <- kappa0 + n
vN <- v0 +n 
vNsigmaN <- v0*sigma0 + (n-1)*s2 + (kappa0*n / (kappa0 + n)) * (ybar - mu0)^2
sigmaN <- vNsigmaN / vN

# Simulations - Gibbs Sampling
sims <- data.frame(mu=0, sigma2=0)
muN <- (kappa0 / (kappa0 + n)) * mu0 + (n / (kappa0 + n)) * ybar 
vNsigmaN <- v0*sigma0 + (n-1)*s2 + (kappa0*n / (kappa0 + n)) * (ybar - muN)^2
sigmaN <- vNsigmaN / vN
X <- rchisq(1, vN)
sigma2 <- (vN * sigmaN / X)
mu <- rnorm(1, muN, sqrt(sigma2/kappaN))  
sims[1,1] <-  mu
sims[1,2] <-  sigma2 
for (i in 2:1000){
  # Byter ut mu0 mot [i-1,1] i sims
  # Byter ut sigma0 mot [i-1,2] i sims
  muN <- (kappa0 / (kappa0 + n)) * sims[i-1,1] + (n / (kappa0 + n)) * ybar 
  # Byter ut mu0 mot muN
  vNsigmaN <- v0*sims[i-1,2] + (n-1)*s2 + (kappa0*n / (kappa0 + n)) * (ybar - muN)^2
  sigmaN <- vNsigmaN / vN
  X <- rchisq(1, n-1)
  sigma2 <- (vN * sigmaN / X)
  mu <- rnorm(1, muN, sqrt(sigma2/kappaN))  
  sims[i,1] <-  mu
  sims[i,2] <-  sigma2 
}
```
The code used to implement the Gibbs sampler that simulates from the joint posterior can be seen in the appendix *R-code*.  

### ii)  

The Gibbs sampler from *i)* is tested and it is of interest to investigate the convergence of the chains. One way to check this is by trace plots.  
```{r, echo=FALSE, fig.height=4, fig.width=8}
#trace plots - with burn-in!
tr_w_burn <- ggplot(sims, aes(x=1:nrow(sims), y=mu)) + geom_line() + theme_bw() + xlab("1:1000") + ylab("Mu") + ggtitle("Gibbs Sampling - Trace Plot - Mu") + geom_vline(xintercept=50, col="red", size=1.05)
tr_w_burn2 <- ggplot(sims, aes(x=1:nrow(sims), y=sigma2)) + geom_line() +
  theme_bw() +ggtitle("Gibbs Sampling - Trace Plot - Sigma^2")+ xlab("1:1000") + geom_vline(xintercept=50, col="red", size=1.05)
grid.arrange(tr_w_burn, tr_w_burn2, ncol=2)
```
By the look of the plots above it seems like the chain has converged for both $\mu$ and $\sigma^2$. The chains converges quickly and if there is a burn-in period, it is thought to be short. Out of the 1000 iterations, perhaps 50 iterations in both cases can be classified as belonging to the burn-in period. Even though it is hard to see a specific burn-in period it is reasonable to make the assumption that some proportion of the first iterations not have converged and should be discarded. 

The chains are thought to have converged since they have a pattern that is...

Next, trace plots for the respective chains shown without the burn-in period.  
```{r, echo=FALSE, fig.height=4, fig.width=8}
tr_wh_burn <- ggplot(sims[51:1000,], aes(x=51:nrow(sims), y=mu)) + geom_line() + theme_bw() + xlab("1:1000") + ylab("Mu") + ggtitle("Gibbs Sampling - Trace Plot - Mu\n Without burn-in")
tr_wh_burn2 <- ggplot(sims[51:1000,], aes(x=51:nrow(sims), y=sigma2)) + geom_line() + theme_bw() + ggtitle("Gibbs Sampling - Trace Plot - Sigma^2\n Without burn-in") + xlab("1:1000")
grid.arrange(tr_wh_burn, tr_wh_burn2, ncol=2)
```

The convergence can also be checked by looking at the autocorrelation of the chains with the burn-in removed.  
```{r, echo=FALSE, fig.height=4, fig.width=8, warning=FALSE}
# Looks at efficieny in terms of auto-correlation
acf_m <- acf(sims[51:1000,1], lag.max = 30, type = c("correlation"), plot=FALSE)
acf_s <- acf(sims[51:1000,2], lag.max = 30, type = c("correlation"), plot=FALSE)
acf_m <- data.frame(ACF=as.numeric(acf_m$acf), Lag=0:30)
acf_s <- data.frame(ACF=as.numeric(acf_s$acf), Lag=0:30)
acf_mu <- ggplot(acf_m, aes(x=Lag, y=ACF))+geom_bar(stat="identity", fill="black")+theme_bw() + ggtitle("Auto-correlation for chain - Mu")
acf_sigma <- ggplot(acf_s, aes(x=Lag, y=ACF))+geom_bar(stat="identity", fill="black")+theme_bw() + ggtitle("Auto-correlation for chain - Sigma^2")
grid.arrange(acf_mu, acf_sigma, ncol=2)
```
Seem to move quite freely, there is no strong correlation between one iteration and the coming iterations. 
Does also speaks for the conclusion that convergence has been reached for both chains. 

## b) Mixture normal model.

### i)
Mattias's code are used for this assignment and added to the appendix at the end of the report. The prior hyperparameters are set to 

Uses standard deviation of observations for assigning prior standard deviation for $\mu$. 
The variance of the observations is used for $\sigma^2_0$
The prior mean is set to 30, close to the mean for the observations (32.27).
For the prior on $\sigma^2$ is the degress of freedom set to 4. 
$\alpha_0$, the prior for the beta distribution, is set (10,10). This is a prior that is used for assigning the probability that a value belong to the distribution. 

```{r, echo=FALSE, results='hide', eval=FALSE}
rainfall <- read.delim("C:/Users/Gustav/Documents/Bayesian-Learning/Lab 3/rainfall.dat",
                       sep="", header = TRUE)
x <- as.matrix(rainfall['X136'])

# Model options
nComp <- 2    # Number of mixture components

# Prior options
alpha <- 10*rep(1,nComp) # Dirichlet(alpha)
alpha <- c(10, 10)
muPrior <- rep(0,nComp) # Prior mean of theta
muPrior <- c(30, 30)
tau2Prior <- rep(sd(x),nComp) # Prior std theta
sigma2_0 <- rep(var(x),nComp) # s20 (best guess of sigma2)
nu0 <- rep(4,nComp) # degrees of freedom for prior on sigma2


# MCMC options
nIter <- 1000 # Number of Gibbs sampling draws


# Plotting options
plotFit <- TRUE
lineColors <- c("blue", "green", "magenta", 'yellow')
################   END USER INPUT ###############

###### Defining a function that simulates from the 
rScaledInvChi2 <- function(n, df, scale){
  return((df*scale)/rchisq(n,df=df))
}

####### Defining a function that simulates from a Dirichlet distribution
rDirichlet <- function(param){
  nCat <- length(param)
  thetaDraws <- matrix(NA,nCat,1)
  for (j in 1:nCat){
    thetaDraws[j] <- rgamma(1,param[j],1)
  }
  thetaDraws = thetaDraws/sum(thetaDraws) # Diving every column of ThetaDraws by the sum of the elements in that column.
  return(thetaDraws)
}

# Simple function that converts between two different representations of the mixture allocation
S2alloc <- function(S){
  n <- dim(S)[1]
  alloc <- rep(0,n)
  for (i in 1:n){
    alloc[i] <- which(S[i,] == 1)
  }
  return(alloc)
}

# Initial value for the MCMC
nObs <- length(x)
S <- t(rmultinom(nObs, size = 1 , prob = rep(1/nComp,nComp))) # nObs-by-nComp matrix with component allocations.
theta <- quantile(x, probs = seq(0,1,length = nComp))
sigma2 <- rep(var(x),nComp)
probObsInComp <- rep(NA, nComp)

# Setting up the plot
xGrid <- seq(min(x)-1*apply(x,2,sd),max(x)+1*apply(x,2,sd),length = 100)
xGridMin <- min(xGrid)
xGridMax <- max(xGrid)
mixDensMean <- rep(0,length(xGrid))
effIterCount <- 0
ylim <- c(0,2*max(hist(x,plot = FALSE)$density))

simulations <- data.frame(w_1 = 0, w_2 = 0, mu_1 = 0, mu_2 = 0)

for (k in 1:nIter){
  alloc <- S2alloc(S) # Just a function that converts between different representations of the group allocations
  nAlloc <- colSums(S)
  # Update components probabilities
  w <- rDirichlet(alpha + nAlloc)
simulations[k,1] <- w[1]
simulations[k,2] <- w[2]  

  # Update theta's
  for (j in 1:nComp){
    precPrior <- 1/tau2Prior[j]
    precData <- nAlloc[j]/sigma2[j]
    precPost <- precPrior + precData
    wPrior <- precPrior/precPost
    muPost <- wPrior*muPrior + (1-wPrior)*mean(x[alloc == j])
    tau2Post <- 1/precPost
    theta[j] <- rnorm(1, mean = muPost, sd = sqrt(tau2Post))
  }
simulations[k,3] <- theta[1]
simulations[k,4] <- theta[2]  
simulations$Expected[k] <- sum(simulations$w_1[k] * simulations$mu_1[k] + 
                                 simulations$w_2[k] * simulations$mu_2[k])
  # Update sigma2's
  for (j in 1:nComp){
    sigma2[j] <- rScaledInvChi2(1, df = nu0[j] + nAlloc[j], scale = (nu0[j]*sigma2_0[j] + sum((x[alloc == j] - theta[j])^2))/(nu0[j] + nAlloc[j]))
  }
  
  # Update allocation
  for (i in 1:nObs){
    for (j in 1:nComp){
      probObsInComp[j] <- w[j]*dnorm(x[i], mean = theta[j], sd = sqrt(sigma2[j]))
    }
    S[i,] <- t(rmultinom(1, size = 1 , prob = probObsInComp/sum(probObsInComp)))
  }
  
  # Printing the fitted density against data histogram
  if (plotFit && (k%%1 ==0)){
    effIterCount <- effIterCount + 1
    #hist(x, breaks = 20, freq = FALSE, xlim = c(xGridMin,xGridMax), main = paste("Iteration number",k), ylim = ylim)
    mixDens <- rep(0,length(xGrid))
    components <- c()
    for (j in 1:nComp){
      compDens <- dnorm(xGrid,theta[j],sd = sqrt(sigma2[j]))
      mixDens <- mixDens + w[j]*compDens
      #lines(xGrid, compDens, type = "l", lwd = 2, col = lineColors[j])
      components[j] <- paste("Component ",j)
    }
    mixDensMean <- ((effIterCount-1)*mixDensMean + mixDens)/effIterCount
    
    #lines(xGrid, mixDens, type = "l", lty = 2, lwd = 3, col = 'red')
    #legend("topleft", box.lty = 1, legend = c("Data histogram",components, 'Mixture'), 
          # col = c("black",lineColors[1:nComp], 'red'), lwd = 2)
  }
}
```


### ii)
The resulting distribution given after 1000 iterations is shown by the plot below.

Picture!
It can be seen the the mixture of the two normals quite well fits the observed data.

The convergence of the Gibbs sampler is analyzed by looking at a trace plot for the chain given by the mixture of the normals.  
```{r, echo=FALSE, fig.height=4, fig.width=8, eval=FALSE}
ggplot(simulations, aes(x=1:nrow(simulations), y=Expected)) + geom_line() + theme_bw() + geom_vline(xintercept=75, col="red", size=1.05) + xlab("1:1000") + ylab("Mu") + ggtitle("Trace plot for chain of values from mixture model")
```
Again, it seems like the chain have converged rather rapidly. Just the first 75 observations are discarded because of the burn-in period. Convergence is thought to have been reached since...

The autocorrelation is also visualised in order to investigate the convergence of the chain.  
```{r, echo=FALSE, fig.height=4, fig.width=8, warning=FALSE, eval=FALSE}
# Looks at efficieny in terms of auto-correlation
acf_mix <- acf(simulations[76:nrow(simulations),5], lag.max = 25, type = c("correlation"), plot=FALSE)
acf_mix <- data.frame(ACF=as.numeric(acf_mix$acf), Lag=0:25)
ggplot(acf_mix, aes(x=Lag, y=ACF))+geom_bar(stat="identity", fill="black")+theme_bw() + ggtitle("Auto-correlation for values from mixture model")
```
The correlation between values from iterations close to each other is low. 

## c) Graphical comparison.

```{r, echo=FALSE, fig.height=4, fig.width=8, eval=FALSE}
## c)
a1 <- data.frame(x=rnorm(1000, 32.27564, sqrt(1546.53868)))
b1 <- data.frame(y=mixDensMean, x=xGrid)
ggplot(rainfall, aes(X136)) + geom_histogram(aes(y = ..density..),alpha=0.9,
  fill="black") + theme_bw() +
  geom_density(data=a1, aes(x),col="royalblue", size=1.05) +
  geom_line(data=b1, aes(x=x, y=y), col="red", size=1.05) + 
  ggtitle("Density for Gibbs Sampling and Mixture of normals\n Blue = 1.a) - Red = 1.b)") + xlab("")
```





# Assignment 2 - Binary regression models

## a)


## b)


## c)

## d)
